---
title: "Logistic Regression Models with R"
author: "Amandine Gamble"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output:
  pdf_document:
    toc: true
    toc_depth: 1
    number_sections: true
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    number_sections: true
urlcolor: blue
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

# Loading Data

We will use the *Palmer Penguin* data set available directly from the `palmerpenguins` package. You can learn more about the data at [allisonhorst.github.io/palmerpenguins](https://allisonhorst.github.io/palmerpenguins/). 

Today, we will explore whether **biometrics can be used to predict sex**.

```{r, results='hide'}

# install.packages("palmerpenguins")
library(palmerpenguins)
library(ggplot2)
library(car)

dat <- as.data.frame(penguins) # rename dataset
str(dat) # structure of dataset

```

Let's focus on one species. 

```{r, results='hide'}

dat_gen = subset(dat, species == "Gentoo" & !is.na(sex))

```

# Exploring the data

*Which one of the variables included in the data may contribute be associated with sex?*

## Plot

```{r, results='hide'}

ggplot(dat_gen, aes(x = flipper_length_mm, y = body_mass_g, color = sex)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("darkorange", "cyan4"), 
                     name = "Sex", 
                     labels = c("Female", "Male", "Unknown")) +
  scale_shape_manual(values = c(16, 17, 15),
                     name = "Species") +
  labs(
    x = "Flipper Length (mm)",
    y = "Body Mass (g)"
  ) +
  theme_minimal() 

```

```{r, results='hide'}

ggplot(dat_gen, aes(x = bill_length_mm, y = bill_depth_mm, color = sex)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("darkorange", "cyan4"), 
                     name = "Sex", 
                     labels = c("Female", "Male", "Unknown")) +
  scale_shape_manual(values = c(16, 17, 15),
                     name = "Species") +
  labs(
    x = "Bill Length (mm)",
    y = "Bill depth (g)"
  ) +
  theme_minimal() 

```

# Logistic Regression to Identify Predictors of Sex

## Re-plot

```{r, results='hide'}

ggplot(dat_gen, aes(x = flipper_length_mm, y = sex, color = sex)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("darkorange", "cyan4"), 
                     name = "Sex", 
                     labels = c("Female", "Male", "Unknown")) +
  scale_shape_manual(values = c(16, 17, 15),
                     name = "Species") +
  labs(
    x = "Flipper Length (mm)",
    y = "Sex"
  ) +
  theme_minimal() 

```

```{r, results='hide'}

ggplot(dat_gen, aes(x = body_mass_g, y = sex, color = sex)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("darkorange", "cyan4"), 
                     name = "Sex", 
                     labels = c("Female", "Male", "Unknown")) +
  scale_shape_manual(values = c(16, 17, 15),
                     name = "Species") +
  labs(
    x = "Weight (g)",
    y = "Sex"
  ) +
  theme_minimal() 

```

```{r, results='hide'}

ggplot(dat_gen, aes(x = bill_length_mm, y = sex, color = sex)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("darkorange", "cyan4"), 
                     name = "Sex", 
                     labels = c("Female", "Male", "Unknown")) +
  scale_shape_manual(values = c(16, 17, 15),
                     name = "Species") +
  labs(
    x = "Bill Length (mm)",
    y = "Sex"
  ) +
  theme_minimal() 

```

```{r, results='hide'}

ggplot(dat_gen, aes(x = bill_depth_mm, y = sex, color = sex)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("darkorange", "cyan4"), 
                     name = "Sex", 
                     labels = c("Female", "Male", "Unknown")) +
  scale_shape_manual(values = c(16, 17, 15),
                     name = "Species") +
  labs(
    x = "Bill Depth (mm)",
    y = "Sex"
  ) +
  theme_minimal() 

```

*Does a linear relationship seem reasonable?*

# Statistical Test

## Assumptions

### Independance

*Observations from each population are independent of one another*. This is likely the case (would be determined by the study design).

### Linearity

*Linearity between the response variable odds and the explanatory variables*. Visible from the scatter plots above.

### Absence of Collinearity

*Absence of collinearity between the explanatory variables*. Suspicious from the scatter plots above.

## Model

#### Model Fitting

```{r, results='hide'}

# Fit model 
mod = glm(sex ~ flipper_length_mm + bill_length_mm + bill_depth_mm + body_mass_g,
          data = dat_gen, 
          family = "binomial")

```

#### Additional Assumption Checks

Double checking for the absence of collinearity.

```{r, results='hide'}

# Variance Inflation Factor
vif(mod)

```

The variance inflation factor is low (< 2) for every variable so we can keep all of them.  

#### Model Selection

```{r, results='hide'}

mod = glm(sex ~ flipper_length_mm,
          data = dat_gen, 
          family = "binomial")
summary(mod)

mod = glm(sex ~ flipper_length_mm + bill_length_mm,
          data = dat_gen, 
          family = "binomial")
summary(mod)

mod = glm(sex ~ flipper_length_mm + bill_length_mm + bill_depth_mm,
          data = dat_gen, 
          family = "binomial")
summary(mod)

mod = glm(sex ~ flipper_length_mm + bill_length_mm + bill_depth_mm + body_mass_g,
          data = dat_gen, 
          family = "binomial")
summary(mod)

```

#### Final Model

```{r, results='hide'}

mod = glm(sex ~ flipper_length_mm + bill_length_mm + bill_depth_mm,
          data = dat_gen, 
          family = "binomial")
summary(mod)

```

#### Coefficient of Determination

```{r, results='hide'}

# Calculate R-squared
deviance <- summary(mod)$deviance
null_deviance <- summary(mod)$null.deviance
rsquared <- 1 - (deviance / null_deviance)
rsquared

```

Based on a logistic regression with forward selection, we found that, if $p$ is the probability of being a male, we can write:

$\text{logit}(p) = \log(\frac{p}{1-p}) = \beta_0 + \beta_1 * flipper + \beta_2 * bill.length + \beta_3 * bill.depth$

With $\beta_0 (SE) = -111.94 (25.40)$, $\beta_1 (SE) = 0.19 (0.10)$, $\beta_2 (SE) = 0.57 (0.22)$ and $\beta_3 (SE) = 2.94 (0.77)$

The $R^2$ correlation coefficient of 0.70, indicates a moderate predictive power (a lot of the variability remains unexplained). 
