---
title: "Linear Regression Models with R"
author: "Amandine Gamble"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output:
  pdf_document:
    toc: true
    toc_depth: 1
    number_sections: true
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    number_sections: true
urlcolor: blue
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

# Loading Data

We will use the *Palmer Penguin* data set available directly from the `palmerpenguins` package. You can learn more about the data at [allisonhorst.github.io/palmerpenguins](https://allisonhorst.github.io/palmerpenguins/). 

Today, we will explore the variability in penguin weight, starting by looking at **whether size and weight are associated**.

```{r, results='hide'}

# install.packages("palmerpenguins")
library(palmerpenguins)
library(ggplot2)
library(dplyr)
library(car)

dat <- as.data.frame(penguins) # rename dataset
str(dat) # structure of dataset

```

# Exploring the data

*Which one of the variables included in the data may contribute be associated with weight?*

## Plot

```{r, results='hide'}

ggplot(dat, aes(x = flipper_length_mm, y = body_mass_g, color = sex, shape = species)) +
  geom_point()

# Improve readability

ggplot(dat, aes(x = flipper_length_mm, y = body_mass_g, color = sex, shape = species)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("darkorange", "cyan4"), 
                     name = "Sex", 
                     labels = c("Female", "Male", "Unknown")) +
  scale_shape_manual(values = c(16, 17, 15),
                     name = "Species") +
  labs(
    x = "Flipper Length (mm)",
    y = "Body Mass (g)"
  ) +
  theme_minimal() 

```

# Associations Between Continuous Variables

## Subset

Let's focus on one species and one sex so we can study intra-species and intra-sex but inter-individual variability. 

```{r, results='hide'}

table(dat$sex, dat$species)

dat_gen_f = subset(dat, species == "Gentoo" & sex == "female") 

```

## Re-plot

```{r, results='hide'}

ggplot(dat_gen_f, aes(x = flipper_length_mm, y = body_mass_g, color = sex, shape = species)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("darkorange", "cyan4"), 
                     name = "Sex", 
                     labels = c("Female", "Male", "Unknown")) +
  scale_shape_manual(values = c(16, 17, 15),
                     name = "Species") +
  labs(
    x = "Flipper Length (mm)",
    y = "Body Mass (g)"
  ) +
  theme_minimal() 

```

*Add regression line*

```{r, results='hide'}

ggplot(dat_gen_f, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm") +
  labs(
    x = "Flipper Length (mm)",
    y = "Body Mass (g)"
  ) +
  theme_minimal() 

```

*Does a linear relationship seem reasonable?*

# Statistical Test

## Assumptions

### Independance (all tests)

*Observations from each population are independent of one another*. This is likely the case (would be determined by the study design).

### Normality (Pearson and regression)

Pearson - *Response variable is normally distributed for each value of the explanatory variable*. This is easier to assess on the model residuals (after model fitting), although we can already guess on the scatter plot. 

```{r}

ggplot(dat_gen_f, aes(x = flipper_length_mm)) +
  geom_histogram() +
  labs(
    x = "Flipper Length (mm)"
  ) +
  theme_minimal() 

ggplot(dat_gen_f, aes(x = body_mass_g)) +
  geom_histogram() +
  labs(
    x = "Body Mass (g)"
  ) +
  theme_minimal() 

```

For a linear regression -  Same as above

### Homoscedasticity (regression)

*Variance of the response variable is equal across each value of the explanatory variable*. This is easier to assess on the model residuals (after model fitting), although we can already guess on the scatter plot. .

## Test

* Null hypothesis: there is no association between flipper length and body weight. 
* Alternative: there is an association between flipper length and body weight (i.e., variations in individual size contributes to explain variations in individual weight).

### Pearson's Correlation Test

Pearson is a parametric test based on the covariance of the two variables. 

```{r}

cor.test(dat_gen_f$flipper_length_mm, dat_gen_f$body_mass_g,
         method = c("pearson"))

```

Based on Pearson's test, we would reject the null hypothesis. There is a significant association between flipper length and body weight (p < 0.05). The Pearson's correlation coefficient [95% confidence interval] of 0.49 [0.26, 0.66], indicates a positive, but weak association (a lot of the variability remains unexplained). 

As the normality assumption was not obviously verified, we can check that we reach the same conclusion with a non-parametric test. 

### Spearman's Rank Test

Spearman is a non-parametric test based on the variables ranks. 

```{r}

cor.test(dat_gen_f$flipper_length_mm, dat_gen_f$body_mass_g,
         method = c("spearman"))

```

Based on Spearman's test, we also reject the null hypothesis. There is a significant association between flipper length and body weight (p < 0.05). The Spearman's correlation coefficient of 0.53, indicates a positive, but weak association (a lot of the variability remains unexplained). 

Both tests support the same conclusion: our conclusion is robust to the violation of the normality assumption.  

*Note that this method does not generate an effect size.*

### Linear Regression Model

Linear models are parametric tests. They can be computed through various methods. A common one is the ordinary least square method.  

#### Model Fitting

```{r}

# Fit model 
mod = lm(body_mass_g ~ flipper_length_mm,
          data = dat_gen_f)

# Return outputs (including p-value)
summary(mod)

```

#### Assumption Checks

Before making any conclusion, we need to check the residuals to ensure the assumptions are verified. This is based on the **model residuals**. 

```{r}

plot(mod)

hist(mod$residuals)

shapiro.test(mod$residuals)

ncvTest(mod)

```

The distribution of the residuals supports the normality and homoscedasticity assumptions. 

#### Conclusion

Based on a linear regression, we reject the null hypothesis. There is a significant association between flipper length and body weight (p < 0.05), with an effect size ± standard error of 35 ± 8 (for every additional mm, a penguin is expected to weight 35 more g). The $R^2$ correlation coefficient of 0.24, indicates weak association (a lot of the variability remains unexplained). 

# Associations Between Continuous and Catergorical Variables

## Subset

Let's now focus on one species to look at the association of weight with sex. 

```{r, results='hide'}

dat_gen = subset(dat, species == "Gentoo") 

```

## Re-plot

```{r, results='hide'}

ggplot(dat_gen, aes(x = sex, y = body_mass_g, color = sex)) +
  geom_jitter(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("darkorange", "cyan4"), 
                     name = "Sex", 
                     labels = c("Female", "Male", "Unknown")) +
  labs(
    x = "Sex",
    y = "Body Mass (g)"
  ) +
  scale_x_discrete(labels = c("Female", "Male", "Unknown")) +
  theme_minimal() 

```

# Statistical Test

We saw before this relationship could be explored with a chi2 or an ANOVA, but they return limited information (notably no effect sizes). We can use linear regressions here too. 

## Assumptions

### Independance

*Observations from each population are independent of one another*. This is likely the case (would be determined by the study design).

### Normality 

For a linear regression - *Response variable is normally distributed for each value of the explanatory variable*. This is easier to assess on the model residuals (after model fitting), although we can already guess on the scatter plot. 

### Homoscedasticity (regression)

*Variance of the response variable is equal across each value of the explanatory variable*. This is easier to assess on the model residuals (after model fitting), although we can already guess on the scatter plot. .

## Test

* Null hypothesis: there is no association between sex and body weight. 
* Alternative: there is an association between sex and body weight (i.e., variations in individual sex contributes to explain variations in individual weight).

### Linear Regression Model

#### Model Fitting

```{r}

# Fit model 
mod = lm(body_mass_g ~ sex,
          data = dat_gen)

# Return outputs (including p-value)
summary(mod)

```

#### Assumption Checks

Before making any conclusion, we need to check the residuals to ensure the assumptions are verified. This is based on the **model residuals**. 

```{r}

plot(mod)

hist(mod$residuals)

shapiro.test(mod$residuals)

ncvTest(mod)

```

The distribution of the residuals supports the normality and homoscedasticity assumptions. 

#### Conclusion

Based on a linear regression, we reject the null hypothesis. There is a significant association between sex and body weight (p < 0.05), with an effect size ± standard error of 805 ± 55 (males are 805 g heavier than females). The $R^2$ correlation coefficient of 0.65, indicates moderate association (some of the variability remains unexplained).

# Multivariate Models

Several explanatory variables can be included in regression models. 

```{r, results='hide'}

ggplot(dat, aes(x = flipper_length_mm, y = body_mass_g, color = sex, shape = species)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("darkorange", "cyan4"), 
                     name = "Sex", 
                     labels = c("Female", "Male", "Unknown")) +
  scale_shape_manual(values = c(16, 17, 15),
                     name = "Species") +
  labs(
    x = "Flipper Length (mm)",
    y = "Body Mass (g)"
  ) +
  theme_minimal() 

```

```{r}

# Fit model 
mod = lm(body_mass_g ~ flipper_length_mm + sex + species,
          data = dat)

# Return outputs (including p-value)
summary(mod)

```

We can see that a large part of the variability is now explained! We will talk more about those models in the coming sessions...