---
title: "R Bootcamp - Part II"
subtitle: "An Introduction to R Programming for Public Health Students"
author: "Amandine Gamble <amandine.gamble@cornell.edu>"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 1
    number_sections: true
urlcolor: blue
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

# Loading Data

We will use the *Number of Reported Malaria Cases by County— United States, 2017* dataset available at [data.cdc.gov/Global-Health/Number-of-Reported-Malaria-Cases-by-County-United-/puzh-5wax/about_data](https://data.cdc.gov/Global-Health/Number-of-Reported-Malaria-Cases-by-County-United-/puzh-5wax/about_data). Download the CSV file onto your computer to be able to easily work with it. 

```{r, results='hide'}

# Load a CSV file
data_malaria = read.csv(
  "Data/Number_of_Reported_Malaria_Cases_by_County__United_States__2017_20250120.csv"
  )
# Notice the changes in the Environmental pane

# Preview the data
head(data_malaria)

```

# Exploring Data (continued)

We can explore and summarize data with functions like `summary()`, `unique()`, `table()`, `aggregate()`...

Data exploration helps understand the dataset structure and key statistics:

```{r, results='hide'}

# Example: Summary statistics
summary(data_malaria)

# Find unique values
unique(data_malaria$STATE)

# Create a contingency table
table(data_malaria$STATE) # Number of observations by state

# Aggregate scores by passing status
aggregate(MAL_FREQ_2017 ~ STATE, data = data_malaria, FUN = sum)

aggregate(MAL_FREQ_2017 ~ STATE, data = data_malaria, FUN = sd)

```

## Activity

  - What does the `aggregate()` function do?
  - Calculate the mean and standard deviation of malaria cases for each state.
  - Create a new data frame containing the information of the number of cases by state. 
  
```{r, include = FALSE}

# Insert answer here

```

# Subsetting Data

Subsetting helps in extracting specific rows, columns, or elements. It can be done with logical tests (`>`, `<`, `==`, `is.na()`) and operators (`!`, `&`, `|`).

```{r, results='hide'}

# Select one element
data_malaria[data_malaria$STATE == "COLORADO",  ]

# Exclude one element
data_malaria[data_malaria$STATE != "NEW YORK",  ]

# Select values above a threshold
data_malaria[data_malaria$MAL_FREQ_2017 > 10,  ]

# Select rows respecting several conditions
data_malaria[data_malaria$STATE == "NEW YORK" & data_malaria$MAL_FREQ_2017 > 10,  ]

```

## Activity

  - What does the `data_malaria[data_malaria$MAL_FREQ_2017 > 10,  ]` command do?
  - Decompose each element of this command:
    - What does `data_malaria$MAL_FREQ_2017` do?
    - What does `data_malaria$MAL_FREQ_2017 < 10` do?
    - What does `data_malaria[...,  ]` do?
  - Subset the data set for the data from Colorado only.
  
```{r, include = FALSE}

# Insert answer here

```

  - What does the `data_malaria[data_malaria$STATE == "NEW YORK" | data_malaria$STATE == "CALIFORNIA",  ]` command do?
  
```{r, include = FALSE}

# Insert answer here

```

# Creating New Variable

New variables can be added to existing data frames with `$variable` or `df[ , "variable"]` (the same way existing variables are called). This is also a way to re-code variables (without erasing the original variable).

```{r, results='hide'}

# Create a new variable
data_malaria$YEAR = 2017

# Attribute values based on conditions
data_malaria[data_malaria$MAL_FREQ_2017 <= 10, "MAL_FREQ_QUAL" ] = "LOW"
data_malaria[data_malaria$MAL_FREQ_2017 > 10, "MAL_FREQ_QUAL" ] = "HIGH"

```

## Activity

  - Re-code malaria frequency as low (10 or less), high (between 10 and 80), very high (80 or more).
  
```{r, include = FALSE}

# Insert answer here

```

# Creating Data Frames from Scratch

vectors can be created `c()`, and combined with `cbind()`, `rbind()`, or `data.frame()`. 

```{r, results='hide'}

# Counties
COUNTY = data_malaria[data_malaria$STATE == "NEW YORK", "COUNTY"]
length(COUNTY)

# Create a new variable
DATA_2018 = c(10, 12, 52, 36, 85, 85, 23, 12, 41, 120, 10, 22)
# Or randomly select numbers
DATA_2019 = sample(length(COUNTY)) 

# Put the data in a data frame
data_malaria_fictive = data.frame(COUNTY = COUNTY,
                                  DATA_2018 = DATA_2018,
                                  DATA_2019 = DATA_2019)

```

## Activity

  - Add DATA_2020 to data_malaria_fictive.
  
```{r, include = FALSE}

# Insert answer here

```

# Pivot 

You can pivot tables for the wide to the long format and vice versa. 

## Activity

  - Pivot data_malaria_fictive from wide to long. You can use AI to help you. 
  
```{r, include = FALSE}

# Insert answer here

```

# Solutions to the activities

```{r, results='hide'}

# Exploring Data ####

# Calculate the mean and standard deviation of malaria cases for each state
aggregate(MAL_FREQ_2017 ~ STATE, data_malaria, mean)
aggregate(MAL_FREQ_2017 ~ STATE, data_malaria, sd)

# Create a new data frame containing the information of the number of cases by state
data_malaria_state = aggregate(MAL_FREQ_2017 ~ STATE, data_malaria, sum)

# Subsetting Data ####

# What does `data_malaria$MAL_FREQ_2017 > 10` do?
data_malaria$MAL_FREQ_2017 > 10
# Test for each county is the number of malaria case is above 10

# What does `data_malaria[...,  ]` do?
data_malaria[data_malaria$MAL_FREQ_2017 > 10, ]
# Call the rows that verify the condition # cases > 10, and all the columns

# Subset the data set for the data from Colorado only.
data_malaria[data_malaria$STATE == "COLORADO", ]

# Creating New Variables ####

# Re-code malaria frequency as low (10 or less), high (between 10 and 80), very high (80 or more).
data_malaria[data_malaria$MAL_FREQ_2017 <= 10, "MAL_FREQ_QUAL" ] = "LOW"
data_malaria[data_malaria$MAL_FREQ_2017 > 10 & 
               data_malaria$MAL_FREQ_2017 < 80, 
             "MAL_FREQ_QUAL" ] = "HIGH"
data_malaria[data_malaria$MAL_FREQ_2017 >= 80, "MAL_FREQ_QUAL" ] = "VERY HIGH"

# Creating Data Frames from Scratch ####

# Add DATA_2020 to data_malaria_fictive.
data_malaria_fictive$DATA_2020 = sample(nrow(data_malaria_fictive))

# Pivot ####

# Pivot data_malaria_fictive from wide to long. You can use AI to help you. 
# Indicate to AI:
# - The language (R)
# - The task (pivot from wide to long)
# - The names of your variables (from colnames(data_malaria_fictive))
# - Adjust the code (e.g., rename object if you need/want)
# - Always check!

colnames(data_malaria_fictive)

# Prompt example: 
# In R, I have a data frame with these variables: "COUNTY"    "DATA_2018" "DATA_2019" "DATA_2020"
# I want to pivot it from wide to long

library(tidyverse)

# Pivot to long format
data_malaria_fictive_long <- data_malaria_fictive %>%
  pivot_longer(cols = starts_with("DATA_"), 
               names_to = "year", 
               values_to = "cases")

# Clean up the year column (remove "DATA_" prefix)
data_malaria_fictive_long <- data_malaria_fictive_long %>%
  mutate(year = str_remove(year, "DATA_"))

print(data_malaria_fictive_long)

```

The End.