---
title: "Comparison of Means across Several Group with R"
author: "Amandine Gamble"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 1
    number_sections: true
urlcolor: blue
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

# Loading Data

We will use the *Palmer Penguin* data set available directly from the `palmerpenguins` package. You can learn more about the data at [allisonhorst.github.io/palmerpenguins](https://allisonhorst.github.io/palmerpenguins/). We will explore how penguin body mass varies across conditions, starting with species. 

```{r, results='hide'}

# install.packages("palmerpenguins")
library(palmerpenguins)
library(ggplot2)
library(dplyr)

dat <- as.data.frame(penguins) # rename dataset
str(dat) # structure of dataset

```

# Exploring the data

## Plot

```{r, results='hide'}

ggplot(dat, aes(x = species, y = body_mass_g, color = sex)) +
  geom_point()

# Improve readability

ggplot(dat, aes(x = species, y = body_mass_g, color = sex)) +
  geom_point(position=position_jitterdodge(dodge.width=0.9)) +
  ylab("Body Mass (g)") +
  xlab("Species") +
  scale_color_discrete(name = "Sex")

```

## Subset

Focus on observations with sex recorded

```{r, results='hide'}

dat = subset(dat, !is.na(sex)) 

unique(dat$sex)

```

## Re-plot

```{r, results='hide'}

ggplot(dat, aes(x = species, y = body_mass_g, color = sex)) +
  geom_point(position=position_jitterdodge(dodge.width=0.9)) +
  ylab("Body Mass (g)") +
  xlab("Species") +
  scale_color_discrete(name = "Sex")

# Display distributions

ggplot(dat, aes(x = body_mass_g)) +
  geom_histogram() +
  xlab("Body Mass (g)") + 
  facet_grid(species ~ sex) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Other option

ggplot(dat) +
  aes(x = species, y = body_mass_g) +
  geom_violin(aes(color = sex), position = position_dodge(0.9)) +
  geom_boxplot(color = "black", aes(fill = sex), width = 0.2, position = position_dodge(0.9))

```

Discuss distribution shapes, and differences (or lack of). 

## Summarize 

```{r, results='hide'}

# mean by group
group_by(dat, sex, species) %>%
  summarise(
    mean = round(mean(body_mass_g, na.rm = TRUE)),
    sd = round(sd(body_mass_g, na.rm = TRUE))
  )

```

# Statistical Test

Let's consider one factor at a time, and consider males only. Here we aim at comparing a continuous variable (body mass) across 3 groups (species). An ANOVA (if assumptions are verified) or a Kruskal-Wallis test would be appropriate.

```{r}

dat_male =  subset(dat, sex == "male")

```

## Assumptions

### Independance 

*Observations from each population are independent of one another*. This is likely the case (would be determined by the study design).

### Normality

*Data is normally distributed within each group*. This is visible on the histogram. Make sure the data is stratified by group.

This can also be verified with the Shapiro-Wilk test, **for each group**.  

```{r}

shapiro.test(dat_male[dat_male$species == "Adelie", "body_mass_g"])
shapiro.test(dat_male[dat_male$species == "Chinstrap", "body_mass_g"])
shapiro.test(dat_male[dat_male$species == "Gentoo", "body_mass_g"])

```

The observed distributions do not statistically differ from a normal distribution. 

### Homoscedasticity

*Variances are equal across populations*. This is visible on the histogram.

This can also be verified with the variance F test test.

```{r}

bartlett.test(body_mass_g ~ species, dat_male) 
```

The variances are not significantly different from one another. 

## Test

```{r}

# Fit model 
mod = aov(body_mass_g ~ species,
          data = dat_male)

# Return outputs (including p-value)
summary(mod)

```

We reject the null hypothesis, i.e., we detect a difference. We cannot tell which group differs from which though. 

## Post-hoc Test

We need to run pair-wise tests with an adjustment for multiple testing to identify which groups differs from which. 

Based on the assumptions we verified above, a t-test is appropriate. 

```{r}

pairwise.t.test(dat_male$body_mass_g, dat_male$species)

```

Gentoos differ from the two others. 

## Explanation

Based on an ANOVA, followed by pair-wise t-tests, male gentoo penguins are significantly heavier than the other species, with gentoo penguins weighting on average (standard deviation) 5,485g (313g) vs 4,043g (347g) for Adelie penguins, and 3,939g (362g) for chinstrap penguins. 

# Non-parametric Equivalent

A non-parametric test would also work (they almost always work!). For a continuous variable recorded across several groups, Kruskal-Wallis rank test is appropriate. 

```{r}

# Kruskal-Wallis test
kruskal.test(body_mass_g ~ species, dat_male)

```

Followed by pair-wise Wilcoxon rank test. 

```{r}

# Wilcoxon test
pairwise.wilcox.test(dat_male$body_mass_g, dat_male$species)

```

We reach the same conclusion. 

# Side note: two-way ANOVAs

We could run a two-way ANOVA to account for sex and species at the same time. 

```{r}

# Fit model 
mod = aov(body_mass_g ~ species + sex,
          data = dat)

# Return outputs (including p-value)
summary(mod)

```

This kind of multivariate models is easier to run and interpret with regression approaches though. 

