---
title: "Multivariate Models with R"
author: "Amandine Gamble"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 1
    number_sections: true
urlcolor: blue
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

# Loading Data

We will use the *Palmer Penguin* data set available directly from the `palmerpenguins` package. You can learn more about the data at [allisonhorst.github.io/palmerpenguins](https://allisonhorst.github.io/palmerpenguins/). 

Today, we will explore the variability in penguin weight.

```{r, results='hide'}

# install.packages("palmerpenguins")
library(palmerpenguins)
library(ggplot2)
library(dplyr)
library(car)

dat <- as.data.frame(penguins) # rename dataset
str(dat) # structure of dataset

```

# Simple Multivariate Models

Several explanatory variables can be included in regression models. 

## Visualization

```{r, results='hide'}

ggplot(dat, aes(x = flipper_length_mm, y = body_mass_g, color = sex, shape = species)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("darkorange", "cyan4"), 
                     name = "Sex", 
                     labels = c("Female", "Male", "Unknown")) +
  scale_shape_manual(values = c(16, 17, 15),
                     name = "Species") +
  labs(
    x = "Flipper Length (mm)",
    y = "Body Mass (g)"
  ) +
  theme_minimal() 

```

## Model

```{r}

# Fit model 
mod = lm(body_mass_g ~ flipper_length_mm + sex + species,
          data = dat)

```

## Assumptions

```{r}

#

```

## Inference

```{r}

# Return outputs (including p-value)
summary(mod)

```

Interpretation...

# More Complexe Multivariate Models

Several explanatory variables can be included in regression models. 

## Visualization

```{r, results='hide'}

ggplot(dat, aes(x = flipper_length_mm, y = body_mass_g, color = sex, shape = species)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("darkorange", "cyan4"), 
                     name = "Sex", 
                     labels = c("Female", "Male", "Unknown")) +
  scale_shape_manual(values = c(16, 17, 15),
                     name = "Species") +
  labs(
    x = "Flipper Length (mm)",
    y = "Body Mass (g)"
  ) +
  theme_minimal() 

```

Let's focus on gentoo penguins

## Subset

```{r}

dat_gep = subset(dat, species == "Gentoo")

```

## Assumptions part 1

### Linearity

```{r}

library(GGally)

ggpairs(dat_gep)

```

## Model

```{r}

# Fit model 
mod = lm(body_mass_g ~ flipper_length_mm + sex + bill_length_mm + bill_depth_mm + year,
          data = dat_gep)

```

## Assumptions part 2

Collinearity

```{r}

# VIF

vif(mod)

```

## Model Selection

```{r}

mod = lm(body_mass_g ~ flipper_length_mm,
          data = dat_gep)

summary(mod)

mod = lm(body_mass_g ~ flipper_length_mm + sex,
          data = dat_gep)

summary(mod)

mod = lm(body_mass_g ~ flipper_length_mm + sex + bill_length_mm,
          data = dat_gep)

summary(mod)

mod = lm(body_mass_g ~ flipper_length_mm + sex + bill_length_mm + bill_depth_mm,
          data = dat_gep)

summary(mod)

mod = lm(body_mass_g ~ flipper_length_mm + sex + bill_length_mm + year,
          data = dat_gep)

summary(mod)

```

Final model

```{r}

mod = lm(body_mass_g ~ flipper_length_mm + sex + bill_length_mm,
          data = dat_gep)

summary(mod)

```

## Assumptions part 3

```{r}

hist(mod$residuals)

plot(mod)

```

## Inference

```{r}

# Return outputs (including p-value)
summary(mod)

```

# Interactions

Here we are looking at the association betwwen weight and 2 categorical variables: sex and species. 

Several explanatory variables can be included in regression models. 

## Visualization

```{r, results='hide'}

ggplot(dat, aes(x = flipper_length_mm, y = body_mass_g, color = sex, shape = species)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("darkorange", "cyan4"), 
                     name = "Sex", 
                     labels = c("Female", "Male", "Unknown")) +
  scale_shape_manual(values = c(16, 17, 15),
                     name = "Species") +
  labs(
    x = "Flipper Length (mm)",
    y = "Body Mass (g)"
  ) +
  theme_minimal() 

```

Let's look at the effect of sex across species

```{r, results='hide'}

ggplot(subset(dat, !is.na(sex)), aes(x = species, y = body_mass_g, color = sex)) +
  geom_boxplot() +
  scale_color_manual(values = c("darkorange", "cyan4"), 
                     name = "Sex", 
                     labels = c("Female", "Male", "Unknown")) +
  labs(
    x = "Species",
    y = "Body Mass (g)"
  ) +
  theme_minimal() 

```

## Model Selection

```{r}

mod = lm(body_mass_g ~ species,
          data = dat)

summary(mod)

mod = lm(body_mass_g ~ species + sex,
          data = dat)

summary(mod)

mod = lm(body_mass_g ~ species + sex + sex:species,
          data = dat)

summary(mod)

```

Final model

```{r}

mod = lm(body_mass_g ~ species + sex + sex:species,
          data = dat)

summary(mod)

```

## Assumptions

```{r}

hist(mod$residuals)

plot(mod)

```

## Inference

```{r}

# Return outputs (including p-value)
summary(mod)

```