---
title: "Comparison of Prevalences across Groups"
author: "Amandine Gamble"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 1
    number_sections: true
urlcolor: blue
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

# Generating the Data

```{r, results='hide'}

library(ggplot2)

```

We will use simulate data on cancer occurrence across smoker and non-smokers. 

```{r, results='hide'}

set.seed(123)

# Parameters
n_total <- 300
prob_smoker <- 0.50           # Overall smoking rate
prob_cancer_given_smoker <- 0.67      # P(Cancer|Smoker) = 100/150
prob_cancer_given_nonsmoker <- 0.133  # P(Cancer|Non-smoker) = 20/150

# Sample all variables using sample()
study_data <- data.frame(
    ID = sample(10000:99999, n_total, replace = FALSE),
  Smoking = sample(c("Smoking", "Non smoking"),    # Sample smoking status
                  n_total, 
                  replace = TRUE, 
                  prob = c(prob_smoker, 1 - prob_smoker))
)

# Sample cancer status based on smoking (vectorized approach)
study_data$Cancer <- ifelse(
  study_data$Smoking == "Smoking",
  sample(c("Cancer", "No cancer"), 
         sum(study_data$Smoking == "Smoking"), 
         replace = TRUE, 
         prob = c(prob_cancer_given_smoker, 1 - prob_cancer_given_smoker)),
  sample(c("Cancer", "No cancer"), 
         sum(study_data$Smoking == "Non smoking"), 
         replace = TRUE, 
         prob = c(prob_cancer_given_nonsmoker, 1 - prob_cancer_given_nonsmoker))
)

# 2. Explore the data
head(study_data)

```

# Visualizing and Summarizing Data

## Contingency Table

*Generate a contingency table of cancer by smoking statuses.*

```{r}

table(study_data$Smoking, study_data$Cancer)

prop.table(table(study_data$Smoking, study_data$Cancer))

addmargins(table(study_data$Smoking, study_data$Cancer))

```

## Probabilities

```{r, results='hide'}

# Smokers
total = sum(study_data$Smoking == "Smoking")
cancer = sum(study_data$Smoking == "Smoking" & study_data$Cancer == "Cancer")
prop.test(cancer, total, conf.level = 0.95)

# Non-smokers
total = sum(study_data$Smoking == "Non smoking")
cancer = sum(study_data$Smoking == "Non smoking" & study_data$Cancer == "Cancer")
prop.test(cancer, total, conf.level = 0.95)

```

The probability (95% confidence interval) of having a cancer is 0.68 [0.61; 0.75] in smokers, and 0.11 [0.07; 0.17] in males. 

## Interpretation

We expect an association between smoking status and cancer status as the confidence intervals for the probability of having a cancer across smoking statuses do not overlap. 

# Hypothesis Testing 

## Biological question

We aim to study the association between smoking status and cancer status. The null hypothesis is that there is no association. 

## Sample and Assumptions

Both variables are recorded as categorical variables. 

```{r, results='hide'}

# Check expected sample sizes
chisq.test(study_data$Smoking, study_data$Cancer)$expected

```

The expected samples sizes in each group under the null hypothesis are > 5, so a chi2 test is appropriate. 

## Inference with chi2 Test

```{r, results='hide'}

# Run test
chisq.test(study_data$Smoking, study_data$Cancer)

```

## Check and Explanation 

A chi2 test rejected the null hypothesis (chi2 = 99.612, p < 0.05), i.e., we detected an association between cancer occurrence and smoking status. The probability (95% confidence interval) of having a cancer is 0.68 [0.61; 0.75] in smokers, and 0.11 [0.07; 0.17] in males. 

# Alternative Tests

## Fisher Exact Test

Appropriate for smaller sample sizes

```{r, results='hide'}

# Run test
fisher.test(study_data$Smoking, study_data$Cancer)

```

Here, we reach the same conclusion as above. 

## `prop.test()`

Uses other existing tests (e.g. chi2)

```{r, results='hide'}

# Run test
# More difficult to use correctly, the successes (Yes) should be 
# 1st in the contingency table
prop.test(table(study_data$Smoking, study_data$Cancer)[, c("Cancer", "No cancer")])

```

Here, it used a chi2, and we reach the same conclusion as above. 
